---
title: "560 Midterm"
author: "Trish Wells"
date: "2/11/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(readxl)
library(tidyverse)
library(caret)
library(FNN)
library(class)
library(rpart)
library(rpart.plot)
```

## #2. B) Impute the missing values in the table.



```{r }
df <- read_excel("C:/Users/wells/Desktop/TBANLT 560/Midterm/impute.xlsx")

df <- df[1:23,]

#Impute Data3 column values using mean
data3nulls <- which(is.na(df$DATA3))

df[data3nulls,]$DATA3 <- mean(df$DATA3, na.rm = TRUE)

#Impute Data5 column using multiple linear regression
data5nulls <- which(is.na(df$DATA5))

lmdata5<-lm(DATA5 ~ DATA1+
              DATA2+
              DATA3+
              DATA4+
              DATA6+
              DATA7+
              DATA8,data=df)
summary(lmdata5) #Only last 3 vars are significant

lmdata5<-lm(DATA5 ~ 
              DATA6+
              DATA7+
              DATA8,data=df)
summary(lmdata5)

b1 <- df[data5nulls,7]*0.36416
b2 <- df[data5nulls,8]*1.09686
b3 <- df[data5nulls,9]*-2.10308

pred <- b1+b2+b3+96.18077
df[data5nulls,6] <- pred
df
```

## #3. A) Split the data into 10 equal (roughly)  partitions.  That means that you will have 633/10 records in each partition.  Determine the number of records with a value of BK = 0 in each partition – how many a value of BK = 0 in each partition.  BK is the variable.

```{r}
df3 <- read_excel("C:/Users/wells/Desktop/TBANLT 560/Midterm/DataforQuestion3.xlsx")


testsplit <- split(df3, f=1:10)
for (i in 1:10){
  assign(paste("part", i, sep=""), as.data.frame(testsplit[i]))
  }


bkvalues <- c(sum(part1$X1.bktype==0),
              sum(part2$X2.bktype==0),
              sum(part3$X3.bktype==0),
              sum(part4$X4.bktype==0),
              sum(part5$X5.bktype==0),
              sum(part6$X6.bktype==0),
              sum(part7$X7.bktype==0),
              sum(part8$X8.bktype==0),
              sum(part9$X9.bktype==0),
              sum(part10$X10.bktype==0))
bkvalues


```
## #3. B)	Generate a subset of the data set with all the records that have a value of BK = 0 (should be 79)  and an equal number of BK=1 (randomly selected from the remaining records, BK=1).    Calculate the mean for DATA1 for each type BK=0, BK=1.


```{r}


bk0 <- df3 %>% filter(bktype==0)
bk1 <- df3 %>% filter(bktype==1)
bk1sample <- bk1[sample(rownames(bk1), nrow(bk0)),]

bk5050 <- rbind(bk0, bk1sample)

DATA1mean <- aggregate(bk5050$DATA1...47, list(bk5050$bktype), FUN=mean)
DATA1mean

```

## #3. C) Conduct a t test for difference in the means of the two groups in the subset.  Put your results here.

```{r}
ttest <-t.test(DATA1...47 ~ bktype, data = bk5050, var.equal = TRUE)
ttest

#p-value <.01 supports alternative hypothesis that true difference in means is not equal to zero; there is a statistical difference between the two means.


```

## #5. A) Calculate the confusion matrix for cutoff of 0.6 and 0.8.

```{r}
library(caret)
ActualY <- c(0,1,0,1,1,0,1)
PredictedY <- c(.5,.9,.7,.7,.3,.4,.5)
confusiondf <- data.frame(ActualY, PredictedY)


matrix.6 <-confusionMatrix(as.factor(ifelse(confusiondf$PredictedY>0.6, '1', '0')), 
                as.factor(confusiondf$ActualY))
matrix.6

matrix.8 <-confusionMatrix(as.factor(ifelse(confusiondf$PredictedY>0.8, '1', '0')), 
                as.factor(confusiondf$ActualY))
matrix.8

confusiondf$Output0.6 <- ifelse(PredictedY>0.6,1,0)
confusiondf$Output0.8 <- ifelse(PredictedY>0.8,1,0)

confusiondf
```


## #5. B) Provide the recall, precision, accuracy, and F1 Measure values for both 0.6 and 0.8.

```{r}
Threshold<- c(0.6,0.8)
error0.6 <- (2+1)/7
error0.8 <- (3+0)/7

accuracy0.6 <- 1-error0.6
accuracy0.8 <- 1-error0.8
Accuracy <- c(accuracy0.6, accuracy0.8)

recall0.6 <- 2/7
recall0.8 <- 1/7
Recall <- c(recall0.6,recall0.8)

precision0.6 <- 2/3
precision0.8 <- 1/1
Precision <- c(precision0.6,precision0.8)

f1.6 <- 2*((precision0.6*recall0.6)/(precision0.6+recall0.6))
f1.8 <- 2*((precision0.8*recall0.8)/(precision0.8+recall0.8))
F1 <- c(f1.6,f1.8)

Accuracytable <- data.frame(Threshold, Accuracy, Recall, Precision, F1)
Accuracytable

```

## #6. Randomly select the data so that there are 10 records for the training data and 10 for the test.  Using KNN identify the test data classes.   Determine the K from possible values of 3, 4, and 5 that will give you the best accuracy for the test cases.   Note: you will want to convert the variable values to integer to perform KNN.


```{r}
df6 <- read_csv("q6_data.csv")
df6$Age <- as.integer(as.factor(df6$Age))

df6$`Spectacle prescription` <- as.integer(as.factor(df6$`Spectacle prescription`))

df6 <- df6 %>% rename(Spec=`Spectacle prescription`, Tear=`Tear production rate`, Class=`Class label Lenses`)

df6$Astigmatic <- as.integer(as.factor(df6$Astigmatic))

df6$Tear <- as.integer(as.factor(df6$Tear))

df6$Class <- as.factor(df6$Class)

trainrows <- sample(row.names(df6), 0.5*dim(df6)[1])
validrows <- setdiff(row.names(df6), trainrows)
traindf <- df6[trainrows,]
validdf <- df6[validrows,]

nn3 <- knn(train=traindf[,2:5], test=validdf[,2:5],
           cl=traindf$Class, k=3)
nn4 <- knn(train=traindf[,2:5], test=validdf[,2:5],
           cl=traindf$Class, k=4)
nn5 <- knn(train=traindf[,2:5], test=validdf[,2:5],
           cl=traindf$Class, k=5)

accuracydf <- data.frame(k=c(3,4,5), accuracy=c(0,0,0))

for(i in 3:5){
  knn.pred <- knn(traindf[,2:5], validdf[,2:5],
                  cl = traindf$Class, k=i)
  accuracydf[i-2,2] <- confusionMatrix(knn.pred,validdf$Class)$overall[1]
}

accuracydf
```


## #7. Construct a decision tree manually using the Gini impurity measure.

```{r}
color <- c("Grey", "Yellow", "Brown", "Grey", "Yellow")
height <- c(10,10,3,10,4)
label <- c("Elephant", "Giraffe", "Monkey", "Elephant", "Tiger")
dtreedf <- data.frame(color, height, label)
EP <- length(which(dtreedf$label == "Elephant"))
GP <- length(which(dtreedf$label == "Giraffe"))
MP <- length(which(dtreedf$label == "Monkey"))
TP <- length(which(dtreedf$label == "Tiger"))
N <- length(dtreedf$label)

gini0 <- 1 - ((EP/N)^2 + (GP/N)^2 + (MP/N)^2 + (TP/N)^2)

# Split by Height, < 10
height0 <- row.names(dtreedf)[which(dtreedf$height<10)]
height10 <- setdiff(row.names(dtreedf), height0)
heightsplit0 <- dtreedf[height0,]
heightsplit10 <- dtreedf[height10,]

# Calculate proportions for heightsplit0
EP0 <- length(which(heightsplit0$label == "Elephant"))
GP0 <- length(which(heightsplit0$label == "Giraffe"))
MP0 <- length(which(heightsplit0$label == "Monkey"))
TP0 <- length(which(heightsplit0$label == "Tiger"))
N0 <- length(heightsplit0$label)

# Calculate proportions for heightsplit10
EP10 <- length(which(heightsplit10$label == "Elephant"))
GP10 <- length(which(heightsplit10$label == "Giraffe"))
MP10 <- length(which(heightsplit10$label == "Monkey"))
TP10 <- length(which(heightsplit10$label == "Tiger"))
N10 <- length(heightsplit10$label)

giniheight0 <- 1 - ((EP0/N0)^2 + (GP0/N0)^2 + (MP0/N0)^2 + (TP0/N0)^2)

giniheight10 <- 1 - ((EP10/N10)^2 + (GP10/N10)^2 + (MP10/N10)^2 + (TP10/N10)^2)

totalginiheight <- (N0/N)*giniheight0 + (N10/N)*giniheight10
# 0.46667

# Now repeat for splitting by color.
# First split tested: Grey vs Yellow/Brown

# Split by Grey vs Yellow/Brown
colorgrey <- row.names(dtreedf)[which(dtreedf$color == "Grey")]
coloryb <- setdiff(row.names(dtreedf), colorgrey)
colorygreysplit <- dtreedf[colorgrey,]
colorybsplit <- dtreedf[coloryb,]

# Calculate proportions for colorgreysplit
EPg <- length(which(colorygreysplit$label == "Elephant"))
GPg <- length(which(colorygreysplit$label == "Giraffe"))
MPg <- length(which(colorygreysplit$label == "Monkey"))
TPg <- length(which(colorygreysplit$label == "Tiger"))
Ng <- length(colorygreysplit$label)

# Calculate proportions for colorybsplit
EPyb <- length(which(colorybsplit$label == "Elephant"))
GPyb <- length(which(colorybsplit$label == "Giraffe"))
MPyb <- length(which(colorybsplit$label == "Monkey"))
TPyb <- length(which(colorybsplit$label == "Tiger"))
Nyb <- length(colorybsplit$label)

ginicolorgrey <- 1 - ((EPg/Ng)^2 + (GPg/Ng)^2 + (MPg/Ng)^2 + (TPg/Ng)^2)

ginicoloryb <- 1 - ((EPyb/Nyb)^2 + (GPyb/Nyb)^2 + (MPyb/Nyb)^2 + (TPyb/Nyb)^2)

totalginicolorgrey <- (Ng/N)*ginicolorgrey + (Nyb/N)*ginicoloryb
# 0.40

# Split by Yellow vs Grey/Brown
coloryellow <- row.names(dtreedf)[which(dtreedf$color == "Yellow")]
colorgb <- setdiff(row.names(dtreedf), coloryellow)
coloryellowsplit <- dtreedf[coloryellow,]
colorgbsplit <- dtreedf[colorgb,]

# Calculate proportions for coloryellowsplit
EPy <- length(which(coloryellowsplit$label == "Elephant"))
GPy <- length(which(coloryellowsplit$label == "Giraffe"))
MPy <- length(which(coloryellowsplit$label == "Monkey"))
TPy <- length(which(coloryellowsplit$label == "Tiger"))
Ny <- length(coloryellowsplit$label)

# Calculate proportions for colorgbsplit
EPgb <- length(which(colorgbsplit$label == "Elephant"))
GPgb <- length(which(colorgbsplit$label == "Giraffe"))
MPgb <- length(which(colorgbsplit$label == "Monkey"))
TPgb <- length(which(colorgbsplit$label == "Tiger"))
Ngb <- length(colorgbsplit$label)

ginicoloryellow <- 1 - ((EPy/Ny)^2 + (GPy/Ny)^2 + (MPy/Ny)^2 + (TPy/Ny)^2)

ginicolorgb <- 1 - ((EPgb/Ngb)^2 + (GPgb/Ngb)^2 + (MPgb/Ngb)^2 + (TPgb/Ngb)^2)

totalginicoloryellow <- (Ny/N)*ginicoloryellow + (Ngb/N)*ginicolorgb
# 0.46667

# Split by Brown vs Yellow/Grey
colorbrown <- row.names(dtreedf)[which(dtreedf$color == "Brown")]
coloryg <- setdiff(row.names(dtreedf), colorbrown)
colorbrownsplit <- dtreedf[colorbrown,]
colorygsplit <- dtreedf[coloryg,]

# Calculate proportions for colorbrownsplit
EPb <- length(which(colorbrownsplit$label == "Elephant"))
GPb <- length(which(colorbrownsplit$label == "Giraffe"))
MPb <- length(which(colorbrownsplit$label == "Monkey"))
TPb <- length(which(colorbrownsplit$label == "Tiger"))
Nb <- length(colorbrownsplit$label)

# Calculate proportions for colorygsplit
EPyg <- length(which(colorygsplit$label == "Elephant"))
GPyg <- length(which(colorygsplit$label == "Giraffe"))
MPyg <- length(which(colorygsplit$label == "Monkey"))
TPyg <- length(which(colorygsplit$label == "Tiger"))
Nyg <- length(colorygsplit$label)

ginicolorbrown <- 1 - ((EPb/Nb)^2 + (GPb/Nb)^2 + (MPb/Nb)^2 + (TPb/Nb)^2)

ginicoloryg <- 1 - ((EPyg/Nyg)^2 + (GPyg/Nyg)^2 + (MPyg/Nyg)^2 + (TPyg/Nyg)^2)

totalginicolorbrown <- (Nb/N)*ginicolorbrown + (Nyg/N)*ginicoloryg
# 0.50

# Splitting by the color Grey gets us the lowest gini.
# If color is grey, label is Elephant. End of branch.

# If color is not grey, we need another branch.
# This branch starts from the colorybsplit table.
# We can either split by height or color.
# Lets test height first. 

# Split by Height, < 10
heightg0 <- row.names(colorybsplit)[which(colorybsplit$height<10)]
heightg10 <- setdiff(row.names(colorybsplit), heightg0)
heightsplitg0 <- colorybsplit[heightg0,]
heightsplitg10 <- colorybsplit[heightg10,]

# Calculate proportions for heightsplitg0
EPg0 <- length(which(heightsplitg0$label == "Elephant"))
GPg0 <- length(which(heightsplitg0$label == "Giraffe"))
MPg0 <- length(which(heightsplitg0$label == "Monkey"))
TPg0 <- length(which(heightsplitg0$label == "Tiger"))
Ng0 <- length(heightsplitg0$label)

# Calculate proportions for heightsplitg10
EPg10 <- length(which(heightsplitg10$label == "Elephant"))
GPg10 <- length(which(heightsplitg10$label == "Giraffe"))
MPg10 <- length(which(heightsplitg10$label == "Monkey"))
TPg10 <- length(which(heightsplitg10$label == "Tiger"))
Ng10 <- length(heightsplitg10$label)

giniheightg0 <- 1 - ((EPg0/Ng0)^2 + (GPg0/Ng0)^2 + (MPg0/Ng0)^2 + (TPg0/Ng0)^2)

giniheightg10 <- 1 - ((EPg10/Ng10)^2 + (GPg10/Ng10)^2 + (MPg10/Ng10)^2 + (TPg10/Ng10)^2)

totalginigheight <- (Ng0/Nyb)*giniheightg0 + (Ng10/Nyb)*giniheightg10
# 0.333333

# Now test split by color, Yellow or Brown
colorgbrown <- row.names(colorybsplit)[which(colorybsplit$color == "Brown")]
colorgyellow <- setdiff(row.names(colorybsplit), colorgbrown)
colorgbrownsplit <- colorybsplit[colorgbrown,]
colorgyellowsplit <- colorybsplit[colorgyellow,]

# Calculate proportions for colorgbrownsplit
EPggb <- length(which(colorgbrownsplit$label == "Elephant"))
GPggb <- length(which(colorgbrownsplit$label == "Giraffe"))
MPggb <- length(which(colorgbrownsplit$label == "Monkey"))
TPggb <- length(which(colorgbrownsplit$label == "Tiger"))
Nggb <- length(colorgbrownsplit$label)

# Calculate proportions for colorgyellowsplit
EPggy <- length(which(colorgyellowsplit$label == "Elephant"))
GPggy <- length(which(colorgyellowsplit$label == "Giraffe"))
MPggy <- length(which(colorgyellowsplit$label == "Monkey"))
TPggy <- length(which(colorgyellowsplit$label == "Tiger"))
Nggy <- length(colorgyellowsplit$label)

ginicolorgbrown <- 1 - ((EPggb/Nggb)^2 + (GPggb/Nggb)^2 + (MPggb/Nggb)^2 + (TPggb/Nggb)^2)

ginicolorgyellow <- 1 - ((EPggy/Nggy)^2 + (GPggy/Nggy)^2 + (MPggy/Nggy)^2 + (TPggy/Nggy)^2)

totalginicolorgbrown <- (Nggb/Nyb)*ginicolorgbrown + (Nggy/Nyb)*ginicolorgyellow
# 0.333333

# The ginis are tied @ 0.3333. Using either split will end up successfully classifying the remaining labels:
## Split by height, then split by Yellow/Brown or
## Split by Yellow/Brown, then split by height.
```

## #8. Say you have 1000 fruits which could be either ‘banana’, ‘orange’ or ‘other’. These are the 3 possible classes of the Y variable. We have data for the following X variables, all of which are binary (1 or 0). If a fruit is long, sweet, and yellow, what is it?

```{r}
Type<-c("Banana", "Orange", "Other", "Total")
Long<-c(400,0,100,500)
NotLong<-c(100,300,100,500)
Sweet<-c(350,150,150,650)
NotSweet<-c(150,150,50,350)
Yellow<-c(450,300,50,800)
NotYellow<-c(50,0,150,200)
Total<-c(500,300,200,1000)
fruitdf<-data.frame(Type,Long,NotLong,Sweet,NotSweet,Yellow,NotYellow,Total)
fruitdf

plong<-fruitdf[4,2]/fruitdf[4,8]
psweet<-fruitdf[4,4]/fruitdf[4,8]
pyellow<-fruitdf[4,6]/fruitdf[4,8]

pbanana<-fruitdf[1,8]/fruitdf[4,8]
porange<-fruitdf[2,8]/fruitdf[4,8]
pother<-fruitdf[3,8]/fruitdf[4,8]

plongbanana<-fruitdf[1,2]/fruitdf[1,8]
psweetbanana<-fruitdf[1,4]/fruitdf[1,8]
pyellowbanana<-fruitdf[1,6]/fruitdf[1,8]

plongorange<-fruitdf[2,2]/fruitdf[2,8]
psweetorange<-fruitdf[2,4]/fruitdf[2,8]
pyelloworange<-fruitdf[2,6]/fruitdf[2,8]

plongother<-fruitdf[3,2]/fruitdf[3,8]
psweetother<-fruitdf[3,4]/fruitdf[3,8]
pyellowother<-fruitdf[3,6]/fruitdf[3,8]

## Probability of being a banana given long, sweet, and yellow
bananaprob <- (plongbanana*psweetbanana*pyellowbanana*pbanana)/
  (plong*psweet*pyellow)

## Probability of being an orange given long, sweet, and yellow
orangeprob <- (plongorange*psweetorange*pyelloworange*porange)

## Probability of being other given long, sweet, and yellow
otherprob <- (plongother*psweetother*pyellowother*pother)/
  (plong*psweet*pyellow)

bananaprob
orangeprob
otherprob
```

